---
title: "Methods and Results"
author: "Simon Garnier"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    css: custom.css
    fig_caption: yes
    number_sections: yes
    theme: cosmo
    toc: yes
---

--- 

```{r, echo=FALSE, message=FALSE, warning=FALSE, 'setup', message=FALSE}
# Load libraries
if (!require(pacman))
  install.packages("pacman")
  
p_load("ggplot2", "cowplot", "dplyr", "RColorBrewer", "viridis")
p_load_gh("swarm-lab/projectSyncedReproduction")

# Create pretty colorblind friendly palette
cbf <- brewer.pal(8, "Dark2")
```

# Methods

## Modelling colony reproductive strategy

The relative number $l$ of larvae in the colony at time $t$ is modeled as a function of 
the form: 

$$ l(t) = (1-A) \left(M-M \left(\frac{\left(\sqrt{10^{2 e}}+1\right) \sin \left(\frac{2 \pi  t}{P}\right)}{\sqrt{10^{2 e} \sin ^2\left(\frac{2 \pi  t}{P}\right)}+1}+1\right)\right)+M \left(\frac{\left(\sqrt{10^{2 e}}+1\right) \sin \left(\frac{2 \pi  t}{P}\right)}{\sqrt{10^{2 e} \sin ^2\left(\frac{2 \pi  t}{P}\right)}+1}+1\right) $$

where $M$ represents the average of the periodical wave, $P$ its period, and $A$ 
its amplitude. The amplitude $A$ is relative to the average $M$ of the periodical 
wave. When $A = 1$ the minimum value of the wave is 0 and the maximum is $2 M$. 
When $A = 0$ the wave is flat (i.e. its minimum and maximum values are both 
equal to the average of the wave). The exponent $e$ control the degree of 
"squarity" of the wave. Positive values of $e$ return a more square-like wave 
while negative values return a more sine-like wave. This allows us to control 
how smooth the reproductive cycle is, in other words, how gradual or abrupt the 
transitions between brood care and reproductive phases are.

For the remainder of this study, we will arbitrarily set the value of $P$, i.e. 
the length of the reproductive cycle, to 1. We will also set the value of $M$, 
i.e. the average relative number of larvae in the colony, to 0.5. As a 
consequence, both the absolute length of the reproductive cycle and the absolute 
number of larvae a colony raises per reproductive cycle are constant across all 
comparisons. With $P = 1$ and $M = 0.5$ we can then simplify the previous equation 
as follows:

$$ l(t) = \frac{A \left(\sqrt{100^e}+1\right) \sin \left(2 \pi  t\right)+\sqrt{100^e \sin ^2\left(2 \pi  t\right)}+1}{2 \left(\sqrt{100^e \sin ^2\left(2 \pi  t\right)}+1\right)} $$

Figure 2 shows the effect of varying either the amplitude $A$, or the "squarity 
exponent" $e$ on the temporal dynamics of the relative number of larvae present 
in the colony across the reproductive cycle.
 
---

```{r, "figure_2", cache=TRUE, echo=FALSE, dpi=300, fig.height=4, fig.width=8}
g1 <- ggplot(data.frame(t = c(0, 3)), aes(t)) +
  stat_function(fun = lt, geom = "line", aes(color = "A = 0; "), size = 1 / 2,
                args = list(A = 0, M = .5, P = 1, e = -10), n = 10000) +
  stat_function(fun = lt, geom = "line", aes(color = "A = 0.5; "), size = 1 / 2 ,
                args = list(A = 0.5, M = .5, P = 1, e = -10), n = 10000) +
  stat_function(fun = lt, geom = "line", aes(color = "A = 1"), size = 1 / 2,
                args = list(A = 1, M = .5, P = 1, e = -10), n = 10000) +
  xlab("Reproductive cycle") + ylab("Relative number of larvae") + 
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", plot.margin = unit(c(0.0, 0.05, 0.0, 0.05), "npc")) +
  scale_color_manual(values = c("A = 0; " = cbf[1], "A = 0.5; " = cbf[2], "A = 1" = cbf[3]), 
                     name = "")

g2 <- ggplot(data.frame(t = c(0, 3)), aes(t)) +
  stat_function(fun = lt, geom = "line", aes(color = "e = 2; "), size = 1 / 2,
                args = list(A = 1, M = .5, P = 1, e = 2), n = 10000) +
  stat_function(fun = lt, geom = "line", aes(color = "e = 1; "), size = 1 / 2,
                args = list(A = 1, M = .5, P = 1, e = 1), n = 10000) +
  stat_function(fun = lt, geom = "line", aes(color = "e = -2"), size = 1 / 2,
                args = list(A = 1, M = .5, P = 1, e = -2), n = 10000) +
  xlab("Reproductive cycle") + ylab("Relative number of larvae") + 
  theme_minimal(base_size = 12) +
  theme(legend.position = "top", plot.margin = unit(c(0.0, 0.05, 0.0, 0.05), "npc")) +
  scale_color_manual(values = c("e = 2; " = cbf[1], "e = 1; " = cbf[2], "e = -2" = cbf[3]), 
                     name = "", breaks = c("e = 2; ", "e = 1; ", "e = -2"))

plot_grid(g1, g2, labels = c("A", "B"))
```
<p class="caption">**Figure 2:** Effect of varying **(A)** the amplitude $A$ or **(B)** the "squarity exponent" $e$ on the temporal dynamics of the relative number of larvae present in the colony across the reproductive cycle.</p>

---

[Back to top]("#")

---

## Modelling colony foraging cost

We consider three possible scenarios for the distribution of foraging costs as a 
function of the number of larvae that have to be fed:

1. **"Proportional":** In this scenario, the cost of foraging grows linearly with 
the number of larvae. This scenario is biologically unlikely but will serve as a 
baseline comparison for the performance of the other two scenarios.

2. **"High Cost of Entry":** In this scenario, the cost of foraging increases 
proportionally faster for smaller numbers of larvae than for larger ones. This 
corresponds to cases where a minimum number of workers are required before 
foraging yields significant benefits (for instance where ants have to overpower 
large prey items or other social insect colonies). This is the scenario likely 
faced by many ant species with army ant-like biology.

3. **"Resource Exhaustion":** In this scenario, the cost of foraging increases 
proportionally slower for smaller numbers of larvae than for larger ones. This 
corresponds to cases where local resources are exploited faster than they are 
replenished, which forces workers to cover increasingly larger foraging 
distances as the number of larvae increases. This is the scenario that is likely 
faced by ant species that mainly forage as scavengers, herbivores, or individual 
predators, i.e. all ant species except those with army ant-like biology.

For all three scenarios, we can model the change in foraging cost $c$ as a 
function of the number larvae $l$ with a function of the form:

$$ c(l) = \frac{1}{2} (n+1) k^{1-n} l^n $$

where $k$ is the maximum number of larvae that a colony can have at any given 
time, and $n$ is a parameter that determines how the cost of foraging scales 
with the number of larvae to be fed. When $n = 1$, the cost of foraging scales 
linearly with the number of larvae ("Proportional" scenario). When $n > 1$, the 
cost of foraging grows slower for smaller than for larger numbers of larvae 
("Resource Exhaustion" scenario). When $0 \leq n < 1$, the cost of foraging 
grows faster for smaller than for larger numbers of larvae ("High Cost of Entry" 
scenario).

Note that this function is designed to ensure that its integral between 0 and 
$k$ is the same regardless of the value of $n$, hence normalizing the foraging 
cost between all possible values of $n$.

For the remainder of this study, we will set $k = 1$, which allows us to 
simplify the previous equation as follows:

$$ c(l) = \frac{1}{2} (n+1) l^n $$

Figure 3 shows the effect of varying $n$ on the shape of the foraging cost 
function. 

---

```{r, "figure_3", echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE, dpi=300, fig.height=4, fig.width=5}
ggplot(data.frame(l = c(0, 1)), aes(l)) +
  stat_function(fun = fc, geom = "line", color = cbf[1], size = 1 / 2,
                args = list(k = 1, n = 3), n = 10000) +
  stat_function(fun = fc, geom = "line", color = cbf[2], size = 1 / 2,
                args = list(k = 1, n = 1), n = 10000) +
  stat_function(fun = fc, geom = "line", color = cbf[3], size = 1 / 2,
                args = list(k = 1, n = 1 / 3), n = 10000) +
  geom_rect(xmin = 1.0025, xmax = Inf, ymin = -Inf, ymax = Inf, fill = "white", color = "white") +
  annotate("text", x = 1.01, y = 2, size = 3, label = "Resource\nExhaustion\nn = 3", hjust = 0, color = cbf[1]) + 
  annotate("text", x = 1.01, y = 1, size = 3, label = "Proportional\nn = 1", hjust = 0, color = cbf[2]) + 
  annotate("text", x = 1.01, y = fc(1, 1, 1 / 3), size = 3,
           label = "High Cost\nof Entry\nn = 1/3", hjust = 0, color = cbf[3]) + 
  scale_x_continuous(breaks = seq(0, 1, 0.25), limits = c(0, 1.2)) + 
  scale_y_continuous(breaks = seq(0, 2, 0.50), limits = c(0, 2.15)) + 
  coord_fixed(ratio = 0.5) +
  xlab("Relative number of larvae") + 
  ylab("Foraging cost") + 
  theme_minimal(base_size = 12) +
  theme(axis.title.x = element_text(hjust = 0.35))
```
<p class="caption">**Figure 3:** Effect of varying the scaling parameter $n$ on the shape of the foraging cost function. The "Proportional" scenario is obtained with $n = 1$. The "High Cost of Entry" scenario is obtained with $0 \leq n < 1$ ($n = 1/3$ is shown here as an example). Finally, the "Resource Exhaustion" scenario is obtained with $n > 1$ ($n = 3$ is shown here as an example). </p>

---

[Back to top]("#")

---

## Integrating reproduction strategy and foraging cost

To evaluate the performance of a given reproductive strategy under different 
foraging cost distributions, we calculate the total foraging cost (i.e. we 
integrate the composite function $c(l(t))$) across one entire colony cycle for 
different values of the relative amplitude $A$ of the reproductive cycle, the 
"squarity exponent" $e$  of the reproductive cycle, and the foraging cost 
scaling parameter $n$. The general shape of the integral function is as follows:

$$ \int c(l(t)) = \frac{1}{2} (n+1) k^{1-n} \int \left((1-A) \left(M-M \left(\frac{\left(\sqrt{10^{2 e}}+1\right) \sin \left(\frac{2 \pi  t}{P}\right)}{\sqrt{10^{2 e} \sin ^2\left(\frac{2 \pi  t}{P}\right)}+1}+1\right)\right)+M \left(\frac{\left(\sqrt{10^{2 e}}+1\right) \sin \left(\frac{2 \pi  t}{P}\right)}{\sqrt{10^{2 e} \sin ^2\left(\frac{2 \pi  t}{P}\right)}+1}+1\right)\right)^n \, dt $$

With $P = 1$, $M = 0.5$ and $k = 1$, we can simplify this equation as follows: 

$$ \int c(l(t)) = 2^{-n-1} (n+1) \int \left(\frac{A \left(\sqrt{100^e}+1\right) \sin \left(2 \pi  t\right)+\sqrt{100^e \sin ^2\left(2 \pi  t\right)}+1}{\sqrt{100^e \sin ^2\left(2 \pi  t\right)}+1}\right)^n \, dt$$

[Back to top]("#")

---

## Software

We used Mathematica 11.0.1.0 to simplify the equations and generate the integral
of the function combining the reproduction strategy with the foraging cost 
$\int c(l(t))$. 

We used the "integrate" function in the "stats" package (version 3.3.2) in R 
(version 3.3.2) to calculate the value of the integral $\int c(l(t))$ over 
various values of the parameters $A$, $e$, and $n$. 

All figures were generated in R using the ggplot2 (version 2.2.0) and cowplot 
(version 0.7.0) packages.

All code used in this manuscript can be found at 
https://github.com/swarm-lab/projectSyncedReproduction

[Back to top]("#")

---

# Results

## Cost of foraging under different foraging scenarios and reproductive strategies

Using the integral function described above, we compute the total cost of 
foraging over a colony cycle for various combinations of the relative amplitude 
$A$ of the reproductive cycle and the shape parameter $n$ of the foraging cost 
function. Since we are not interested here in the effect of the shape of the 
reproductive cycle, which will be treated in the following section, we set its 
shape to a near-square wave ($e = 10$). Note, however, that the results are 
qualitatively equivalent with a sine wave. Figure 4 summarizes the results.

---

```{r, message=FALSE, warning=FALSE, "data_figure_4", cache=TRUE, include=FALSE}
A <- matrix(rep(seq(0, 1, length.out = 201), 201), 201, 201)
n <- matrix(rep(10 ^ seq(-log10(4), log10(4), length.out = 201), each = 201), 201, 201)
n_log10 <- log10(n)

dat <- data.frame(A = c(A), n = c(n)) %>%
  group_by(A, n) %>%
  do(total_cost = integrate(tc, 0, 1, 
                            M = 0.5, A = .$A, P = 1, e = 10,
                            k = 1, n = .$n)$value) %>%
  mutate(total_cost = unlist(total_cost),
         n_log10 = log10(n)) %>%
  ungroup()

m <- matrix(log(dat$total_cost), 201, 201)
g <- approxGradient(m)

A_small <- t(A)[seq(5, 201, 5), ]
A_small <- A_small[, seq(5, 201, 5)]
n_small <- t(n_log10)[seq(5, 201, 5), ]
n_small <- n_small[, seq(5, 201, 5)]

r <- g$r / sqrt(g$r ^ 2 + g$c ^ 2)
c <- g$c / sqrt(g$r ^ 2 + g$c ^ 2)

x_end <- t(A) - r / 40
x_end <- x_end[seq(5, 201, 5), ]
x_end <- x_end[, seq(5, 201, 5)]

y_end <- t(n_log10) - c / 40
y_end <- y_end[seq(5, 201, 5), ]
y_end <- y_end[, seq(5, 201, 5)]

gradient_dat1 <- data.frame(x = c(A_small), y = c(n_small), 
                           x_end = c(x_end), y_end = c(y_end))
```

```{r, "figure_4", cache=TRUE, echo=FALSE, dpi=300, fig.height=4, fig.width=6}
ggplot() +
  geom_tile(data = dat, aes(x = A, y = n_log10, fill = log(total_cost))) + 
  stat_contour(data = dat, aes(x = A, y = n_log10, z = log(total_cost)), 
               color = "#FFFFFF50", bins = 25) +
  geom_hline(yintercept = 0, color = "#FFFFFF", linetype = 2) +
  geom_segment(data = gradient_dat1,
               aes(x = x, y = y, xend = x_end, yend = y_end),
               arrow = arrow(length = unit(0.005, "npc")), 
               color = "#FFFFFF50") +
  scale_y_continuous(breaks = log10(c(1/4, 1/3, 1/2, 3/4, 1, 4/3, 2, 3, 4)),
                     labels = c("1/4", "1/3", "1/2", "3/4", "1", "4/3", "2", "3", "4")) +
  coord_cartesian(xlim = c(0, 1), ylim = c(-0.61, 0.61)) +
  xlab("Relative amplitude A") + ylab("Shape parameter n") +
  scale_fill_viridis(name = "log(total cost)") + 
  theme_minimal(base_size = 12)
```
<p class="caption">**Figure 4:** Total foraging cost over a reproductive cycle 
for various combinations of the relative amplitude $A$ of the reproductive cycle 
and the shape parameter $n$ of the foraging cost function. Values of $n$ above 
the dashed line (which indicates the “Proportional” scenario) correspond to 
“Resource Exhaustion” scenarios, while values below the dashed line correspond 
to “High Cost of Entry” scenarios. Isolines (white contour lines) represent 
points in the parameter space with constant value, and gradient vectors (white 
arrows) represent the direction, but not the intensity of the local gradient.</p>

---

When the foraging cost function corresponds to a “Resource Exhaustion” scenario 
($n > 1$), the lowest total foraging cost is obtained by a non-phasic 
reproductive strategy that distributes the number of larvae produced by the 
colony uniformly in time (i.e. $A = 0$; see top left part of Figure 4).

On the contrary, when the foraging cost function corresponds to a “High Cost of 
Entry” scenario ($0 \leq n < 1$), the total foraging cost decreases with 
periodical variation in the number of larvae produced by the colony. Lower costs 
are achieved for larger oscillation amplitudes, and colonies perform best under 
the most extreme phasic reproductive strategy in terms of oscillation amplitude 
(i.e. $A = 1$; see bottom right part of Figure 4).

---

[Back to top]("#")

---

## Effect of smoothness of phase transitions

Here we test the effect of abrupt phase transitions in which all larvae hatch 
and pupate at the exact same time (i.e. a square wave cycle) versus smooth phase 
transitions in which larvae hatch and pupate around an average time (i.e. a sine 
wave cycle) in a "High Cost of Entry" scenario. We do not test this effect in a 
“Resource Exhaustion” scenario because the results in the previous section show 
that a perfectly non-phasic reproductive strategy is favored in this case.

Using the integral function described above, we compute the total cost of 
foraging over a reproductive cycle for different values of the cycle’s "squarity 
exponent" $e$. We set $A$ to 1, i.e. a cycle with maximum oscillation intensity, 
and $n$ to 1/4, i.e. a "High Cost of Entry" scenario. Note that results are 
qualitatively similar for any combination of $A > 0$ and $0 < n < 1$. Figure 5 
summarizes the results.

---

```{r, message=FALSE, warning=FALSE, "data_figure_5", cache=TRUE, include=FALSE}
dat <- expand.grid(e = seq(-5, 10, length.out = 1000)) %>%
  group_by(e) %>%
  do(total_cost = integrate(tc, 0, 1, 
                            M = 0.5, A = 1, P = 1, e = .$e, 
                            k = 1, n = 1 / 4)$value) %>%
  mutate(total_cost = unlist(total_cost))
```  

```{r, echo=FALSE, fig.height=4, fig.width=5, "figure_5", cache=TRUE, dpi=300}
ggplot(dat, aes(x = e, y = total_cost)) + 
  geom_path(size = 1) + 
  xlab("Squarity exponent e") + ylab("Total foraging cost") +
  theme_minimal(base_size = 12) 
```
<p class="caption">**Figure 5:** Effect of the smoothness of phase transitions 
(determined by the “squarity exponent” $e$) on the total foraging cost over a 
reproductive cycle for colonies in a "High Cost of Entry" foraging scenario.</p>

---

The total cost of foraging decreases with the value of the "squarity exponent" 
$e$, indicating that abrupt phase transitions are more beneficial than smooth 
phase transitions for colonies experiencing a "High Cost of Entry" foraging 
scenario.

[Back to top]("#")

---


